---
title: "Session Five"
author: "Akos Mate"
subtitle: "Writing functions and iterating in R"
date: '2018 July'
output:
    html_document:
        toc: true
        toc_depth: 3
        theme: readable
        css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE
)
```

# 1. Loops

This session focuses on ways to save time and keystrokes: writing functions and iterations. As a rule of thumb if you find yourself writing (copying) the same code more than twice then it should be a function (or loop). This section will be more focused on base R solutions, so we will write loops, then embed our loops in functions. Then, we will discuss two equivalent to writing loops: the `apply` function family from base R, and the `purrr:map` family from the `tidyverse`.
  
  
Loops are a way of iterating a given operation over a set of different inputs. We start by loading our packages and a subsetted msleep data. 

```{r, echo=FALSE}
# !diagnostics off
```


```{r}
library(dplyr)
library(purrr)
library(ggplot2)
```

```{r}
msleep_df <- msleep %>% 
    select(name, sleep_total, sleep_rem, awake)

set.seed(2018) # for reproducability
```

## 1.1 For loops

A for loop looks like:

```{r eval=FALSE}
for (value in that) {
this
}
```

What a for loop essentially does is that for every `value` in the `that` vector/object, do the `this` operation. It's best to explore this via a very simple excercise. The following  script gives you a very simple code, where we take every element (`i`) (but it could be named anything really) in the `1:5` vector and have the `print()` function print out the values.


```{r}
# simple for loop
for (i in 1:5) {
    print(i)
}

1:5



# note the different output!
 


```

We can set up a loop to perform a set of operations on our input vector and put the result in a pre-specified output vector. Let's take the msleep data and compute the mean value for each column.

```{r error=TRUE, warning=TRUE}
msleep_num <- msleep_df %>%
    select_if(is.numeric)

ms_colmean <- vector("double", ncol(msleep_num)) # if we know our output lenght we should create a vector with that lenght.

for (i in seq_along(msleep_num)) {
    ms_colmean[i] <- mean(msleep_num[i])
}

```


Why did we get an error? (think about the difference between `[` and `[[`)

hint:
```{r}
msleep_num[2] 

# or

msleep_num[[2]]
```


```{r}
for (i in seq_along(msleep_num)) {
    ms_colmean[[i]] <- mean(msleep_num[[i]])
    print(ms_colmean[i])
}


# to get rid of the NA, use the `na.rm = TRUE` argument.

for (i in seq_along(msleep_num)) {
    ms_colmean[[i]] <- mean(msleep_num[[i]], na.rm = TRUE)
    print(ms_colmean[i])
}
```

We used the `seq_along` funtion to define our loop sequence (previously we just went with `1:whatever`). Another common way to define the sequence is the `1:lenght(input)`. In the off chance that you create a vector with 0 lenght however, `seq_along` will display a correct output, while `1:lenght(input)` won't.

```{r}
zero <- vector("double", 0)

1:length(zero)

# vs

seq_along(zero)
```

Loops can be nested in each other as well. To demonstrate this, we will do a multiplication table (a `10x10` matrix) with a nested for loop.

```{r}

# nested loop for a multiplication table
# right now it is populated by NA's
(mult_table <- matrix(NA, nrow = 10, ncol = 10))

num1 <- 1:10 # our input vector

for (i in num1) {               
    for (j in num1) {           
        mult_table[i,j] <- i*j 
    }
}

mult_table

```

## 1.2 conditions (if, else)

It is a conditional statement, which we can put in a for loop if we want.

```{r eval=FALSE}
if (this) {
Plan A
} else {
Plan B
}

```

A quick example(s).

```{r}
x <- 5

if (x > 5) {
    print("This is greater than 5")
} else {
    print("this not as great as 5 :(")
}


# a more complicated example using multiple if conditions, where we are curious if the input is even or odd. We also want a nicer output with more communication.

input <- c(1:6)

for (i in seq_along(input)){
     stop_cond <- is.integer(input)
     if (!stop_cond) {
         stop("input must be integer!") # given the nature of our test condition, this only works on integers
     } else {
         num_test <- input[i]%%2 == 0 # the `%%` operator returns the modulus. if 0 then even, if not, odd.
         if (num_test) {
             cat(i, "is even; ") # cat is a more flexible print function, which can combine objects and strings
         } else {
             cat(i, "is odd; ")
         }
     }
 }
```

A more practical example on using conditional statement is to recode a variable. We will create a new column in our curtailed msleep_df data frame and fill it with `NA`'s initially. Then we will create a dummy variable, which is 1 if `sleep_value > 1` AND `awake_value > 18` and 0 if this condition is not met.

```{r}
# if else condition inside the loop
msleep_df$new_awake <- NA

for (i in 1:nrow(msleep)) {
    sleep_value <- msleep_df$sleep_total[i]
    awake_value <- msleep_df$awake[i]
    
    test <- sleep_value > 1 & awake_value > 18
    
    if (test) {
        msleep_df$new_awake[i] <- 1
    } else {
        msleep_df$new_awake[i] <- 0
    }
}

msleep_df$new_awake
```


> Excercise: recode gapminder gdp variable to below and above average dummy with a loop. The result should be something similar to below.

solution:
```{r}
gapminder <- gapminder::gapminder

gapminder$gdp <- NA
gdp_mean <- mean(gapminder$gdpPercap, na.rm = TRUE)

for (i in 1:nrow(gapminder)) {
    gdp_test <- gapminder$gdpPercap[i] > gdp_mean
    
    if (gdp_test) {
        gapminder$gdp[i] <- "above"
    } else {
        gapminder$gdp[i] <- "below"
    }
}


# let's double check if our loop works correctly. (don't mind the stringr package for now)
sum((gapminder$gdpPercap) < mean(gapminder$gdpPercap)) == sum(stringr::str_count(gapminder$gdp, "below"))

```

```{r,}
head(gapminder, 10)
```

## 1.3 While loop

```{r eval=FALSE}
while (condition){
  # Do whatever is here as long as the condition is TRUE. In each iteration of the loop, the condition much be updated according to a certain logic, and it is evaluated again at the beginning of the loop to decide whether to go through with the next iteration or to stop the loop. 
}
```


For illustrative purposes, let's rewrite our first little for loop!


```{r eval=FALSE}
i <- 0 # set our initial value

while (i < 5 ) {
    print(i)
}
```

Press the `Esc` to stop our infite loop! What just happened? We need to ensure that at one point, our while condition is met so our loop ends.

```{r}

while (i < 5 ) {
    print(i)
    i <- i+1 # this adds +1 to our `i` which then will reach 5 and stop our loop.
}
```


# 2. Functions

For functions the same logic applies: if you have to copy paste/write the same line twice, think of a way to turn it into a function. The syntax of the `funtion()` function is the following:

```{r eval=FALSE}
name <- function(variables) {
    this is where we define our function. 
}
```

As with loops, you need to be consistent within your function with the namings of various interim objects, inputs and outputs. To get a feel for creating a function, let's create one, which will exponentiate a choosen base to our choosen exponent.

```{r}
my_power <- function(base, exp){
    output <- base ^ exp
    return(output)
}

# check out our function
my_power(base = 2, exp = 6)
```

Conditional statements within the functions work according to the same logic as in the loops discussed previously. We should add a much needed error message to our function:

```{r}
# add some error messages to our function with an if else + stop combination
my_power2 <- function(base, exp){
    cond <- is.numeric(base)
    if (!cond) {
        stop("base must be numeric!")
    } else {
        output <- base ^ exp
        return(output)
    }
}
```

```{r error=TRUE, collapse=FALSE}
my_power2("2", 4)

# we can experiment, as long as our inputs are numeric:
my_power2(2, 1:5)

my_power2(1:5, 2)
```

Or we can simulate a dice roll, with the use of the `sample()` function. If you are interested in how to build such simulations in R, you can check out _Grolemund, Garrett. Hands-On Programming with R: Write Your Own Functions and Simulations_

```{r}
# create a dice rolling function
roll <- function(){
    die <- 1:6
    dice <- sample(die, size = 1, replace = TRUE)
    
    return(dice)
}

roll()
```


> **Quick excercise:** write a function, which standardizes (creates z scores from raw scores) an input vector. The formula for the standardization is: $$z=\frac{x-\bar{x}}{S}$$  
> where, $x$ is the raw score (numeric value in our input vector) in our sample; $\bar{x}$ is the sample mean; and $S$ is the sample standard deviation.  


Solution:
```{r}
z_score <- function(x, sample) {
    
    output <- (x-mean(sample, na.rm = TRUE))/sd(sample, na.rm = TRUE)
    
    return(output)
}

# let's check our function with some random, normally distributed data
height <- rnorm(50, 0, 1)

z_score(height[1:5], height) # it works! 

```



It is now time to put our knowledge to good practical use and combine a function and loops. Since we calculate column means frequently, we should just probably write a function for it.

```{r}
column_mean <- function(df) {
    output <- vector("double", length(df))
    for (i in seq_along(df)) {
        output[i] <- round(mean(df[[i]]),2)
    }
    
    return(output)
}

column_mean(msleep_df)
```

If we want to have a more general summary function, we can supply a function as argument to our function (Xzibit would be so proud!)

```{r collapse=FALSE}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  return(out)
}

col_summary(msleep_df, sd)

col_summary(msleep_df, median)

```


# 3. apply function family

There are several arguments against loops in the R community. They used to be slow (that changed in recent years), so people preferred using the vectorized solution of the `apply` family. There are many task that you can use vectorized operations in R. For example:

```{r collapse=FALSE}
input2 <- c(1:4)

for (i in seq_along(input2)) {
    print(i+2)
}



# or

(input2 + 2) # if we put the whole operation in () it will print by default after execution.

```

The `apply` family functions are usually very fast, so if you are dealing with large data, they can save considerable time. In this section we'll go over `apply()`, `lapply()`, 

## 3.1 `apply`

The `apply` function let's us apply a function to the rows or columns of our data frame or matrix by adjusting the `MARGIN = ` argument. 1 for row, 2 for column

```{r}
df <- data.frame(x = rnorm(5),
                 y = rnorm(5),
                 z = rnorm(5))

df
```

```{r}
# sum over each column
apply(df, 2, sum)

# sum over each row
apply(df, 1, sum)

```

We can use it to check the number of missing values in our data frame as well (which is a very useful thing to do). here we "wrap" our function with the `function(x)`, otherwise we'll get an error. When in doubt, you can add `function(x)` even when it is redundant it won't cause any problem

```{r}
apply(msleep_num, 2, function(x) sum(is.na(x)))

# redundant `function(x)`
apply(df, 2, function(x) sum(x))


```


> Quick excercise: Let's calculate the column means of the msleep_num data frame. Use the `apply()` function!  
> you should get something similar:  

```{r}
apply(msleep_num, 2, function(x) mean(x, na.rm = TRUE))

```

## 3.2 `lapply`

The `lapply` function slightly differs from the `apply`:  
* It takes two arguments: `lapply(list, function)`
* It iterates the function over vectors or lists. This means that our output will also be a list.

Let's see what happens when we put a data frame into `lapply`. As a data frame is essentially lists (as columns) put together we get a result for each column (as a list). If we want a vector, we need to embed our `lapply()` function in an `unlist()`.
```{r}
# output as list
lapply(df, sum)

# output as vector
unlist(lapply(df, sum))
```

We can use it to create a list where every element is a matrix.

```{r}
mat_out <- lapply(1:3, function(x) matrix(x, nrow = 5, ncol = 5))

mat_out
```


## 3.3 `sapply`
```{r}
sapply(df, sum)

sapply(df, sum, simplify = FALSE)

sapply(msleep_num, function(x) mean(x, na.rm = TRUE))
```


# 4. `purrr` package for iteration

Another way of iterating over our data is the `map` function from the `purrr` package. You can specify what sort of output you want after the `map_` part.

```{r}
map_dbl(df, mean)

map_dbl(mat_out, mean)

msleep_df %>% 
    select_if(is.numeric) %>% 
    map_dbl(mean, na.rm = TRUE)
```


